---
title: "Dataset Tidying and Preparation"
author: "Jack Goldman"
date: "2023-01-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load required packages

```{r}
library(tidyverse)
library(sf)
library(raster)
library(ggplot2)
library(tmap)
library(exactextractr)
library(kableExtra)
```

### read in data

```{r}
#snow disappearance date
sdd_path <- "C:/Users/jandr/Documents/PhD/Thesis/Data/SDD/"
sdd_list <- list.files(sdd_path, pattern = glob2rx("*Shield.tif"))

#snow cover frequency
scf_path <- "C:/Users/jandr/Documents/PhD/Thesis/Data/SCF/"
scf_list <- list.files(scf_path, pattern = glob2rx("*Shield.tif"))

# Ontario Secondary Watersheds
ws = st_read("C:/Users/jandr/Documents/PhD/Thesis/Data/Watersheds/ONT_WSHED_BDRY_SEC_DERIVED.shp")

#Ontario Shield
shield = st_read("C:/Users/jandr/Documents/PhD/Thesis/OntBSdb/Watersheds_Ecoregion/Ontario_Shield_v0.shp")
```

### Visualize watersheds

```{r, viz watersheds, cache = TRUE}
tm_shape(ws) +
  tm_fill() +
  tm_borders() 

```

### Clip watersheds to the ontario shield ecoregion

```{r, clip watersheds to shield, cache = TRUE}
# transform crs between both objects to match
ws_crs = st_crs(ws) # EPSG 4269
shield_crs = st_crs(shield) # EPSG 4326
#transform to 4326 -- WGS 84
ws = ws %>%  st_transform(shield_crs)
#check transformation to confirm matching
st_crs(ws) == st_crs(shield)

#crop watershed by shield geometry
ws_shield  = st_intersection(ws, shield)

#lets map new ws to check 
wss = tm_shape(ws_shield) +
      tm_fill() +
      tm_borders() 
wss
```

## Now lets plot read snow disappearance date in 2001

There are a few steps we have to take. Step 1: read in raster and crop to area of interest

```{r, cache = TRUE}
#read in raster -- select first raster in list which is 2001
sdd2001 = raster::raster(paste0(sdd_path, sdd_list[1]))

#lets crop sdd2001 to shield -- first check crs
st_crs(sdd2001) == st_crs(shield) # all good

#crop
sdd2001 = crop(sdd2001, shield)
sdd2001 = raster::mask(sdd2001, shield)

```

Step 2: Lets look at the distribution of values in the rasters This will help us set a palette to better visualize the raster later on

```{r, cache = TRUE}
sdd2001 %>% as.data.frame() %>% 
  ggplot(aes(x = SDD))+
  geom_density(alpha = 0.3)

# lets get a tibble of summary statistics

sdd2001 %>% 
  as.data.frame() %>% 
  filter (SDD > 0) %>% # we not want 0 which is permanent or waterbody
  summarise(mean = mean(SDD),
            max = max(SDD),
            min = min(SDD),
            q25 = quantile(SDD, 0.25),
            q75 = quantile(SDD, 0.75))
```

Step 3: Visualize the raster using custom breaks

```{r, cache = TRUE}
#lets set custom breaks based on summary values
breaks_01 = c(0, 94, 214, 220, 245, 265,  297)

# plot
wss1 = wss + 
  tm_shape(sdd2001) +
  tm_raster(alpha = 0.7, palette = "Blues", breaks = breaks_01)

wss1
```

### now lets calculate the average for each watershed

```{r, results= FALSE, cache = TRUE}

ws_shield = 
  ws_shield %>% mutate(
    sddMean_01 = exact_extract(sdd2001, ws_shield, 'mean'),
    sddMax_01 = exact_extract(sdd2001, ws_shield, 'max')
  )

```

Lets View it in a table
```{r}
ws_shield %>% 
  st_set_geometry(NULL) %>%
  knitr::kable() %>% 
  kable_styling(font_size = 8)
```



Lets plot it

```{r, cache = TRUE}
wss_sddmean_01 = 
  tm_shape(ws_shield) +
  tm_polygons(col = 'sddMean_01', palette = "Blues" )

wss_sddmean_01
```
### lets bring in fires and plot the fire locations for the year 2001

```{r}
fires = st_read("C:/Users/jandr/Documents/PhD/Thesis/OntBSdb/On_Fire_Perimeters01to2020.shp")
```

Get 2001 fire perimeters and calculate centroids
```{r, cache=TRUE}
fires_01 = fires %>% 
  filter(Fire_Year == 2001) %>% 
  st_centroid()

```
Plot them
```{r, plot fire points, cache = TRUE}
wss_sddmean_01_fp = 
  wss_sddmean_01 + 
  tm_shape(fires_01) +
  tm_dots(size = 0.1, col = "red" , shape = 4)

wss_sddmean_01_fp
```


plot by size
```{r}
wss_sddmean_01_fs = 
  wss_sddmean_01 + 
  tm_shape(fires_01) +
  tm_symbols(col = "FIRE_FINAL", size = 0.1)

wss_sddmean_01_fs
```


Get area burned per watershed
```{r}

ws_shield_f = st_join(fires_01, ws_shield)

ws_shield_f = 
  ws_shield_f %>% 
  group_by(NAME) %>% 
  summarise(area_burned = sum(FIRE_FINAL)) %>% 
  st_drop_geometry()

ws_shield = 
  ws_shield %>% 
  left_join(ws_shield_f, by = "NAME") %>% 
  replace_na(list(area_burned = 0))

```

plot area burned
```{r}
wss_areaburned = 
  tm_shape(ws_shield) +
  tm_polygons(col = "area_burned", palette = "Reds" )

wss_areaburned
```

Lets plot the two maps side by side
```{r}
tmap_arrange(wss_sddmean_01, wss_areaburned)

```

# lets look at defoliation

```{r}
defol_01 = st_read("C:/Users/jandr/Documents/PhD/Thesis/Data/Defoliation/SBW_ON_1984-2020_CR/2001_ON_2022_02_10_CR.shp")

defol_01 %>%
  st_set_geometry(NULL) %>% 
  glimpse()
```
crop it to study area
```{r}
# check crs
st_crs(defol_01) == st_crs(ws_shield)

#clip

defol_01 = 
  defol_01 %>% 
  st_intersection(ws_shield)
```
```{r}
wss_defol = 
  wss +
  tm_shape(defol_01) +
  tm_polygons(col = "DEFOL", palette = "Greens")
wss_defol

```

